---
title: "Bacterial communities associated with distinct coral-*Symbiodinium* holobionts in St. John, U.S. Virgin Islands"
author: "Ross Cunning"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document: 
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: true
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = normalizePath(".."))
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, cache = TRUE)
options(scipen = 3, digits = 9)
```

```{r setup_2, include=FALSE}
# Load custom R functions
source("R/functions.R", .GlobalEnv)
# Load package libraries
library(phyloseq); library(vegan); library(multcompView); library(reshape2); library(igraph); library(stringr); library(tidyverse); library(cowplot); library(readxl)
# Set colors for plotting clades
taxcolors <- matrix(c("#8dd3c7", "#ffffb3", "#bebada", "#fb8072", "#80b1d3", "#fdb462"), 
                    dimnames=list(c("CladeA", "CladeB", "CladeC", "CladeD", "CladeF", "CladeG")))
```

# Import data

### Load *Symbiodinium* data
```{r}
load("data/phy.f.p.RData")
sym <- phy.f.p

# Remove samples that were run twice and rename
sym <- subset_samples(sym, sample_names(sym) != "LT071A")
sym <- subset_samples(sym, sample_names(sym) != "OC084A")
sample_names(sym)[sample_names(sym) == "LT071B"] <- "LT071"
sample_names(sym)[sample_names(sym) == "OC084B"] <- "OC084"

# Change order of samples in phyloseq, because distance matrix depends on this
samplenames <- sort(sample_names(sym))
sym_samdat <- sample_data(sym) %>%
  arrange(InputFileName) %>%
  mutate(sample = samplenames) %>%
  column_to_rownames("sample") %>%
  data.frame() %>%
  sample_data()
sym_otutab <- as.matrix(otu_table(sym))[, samplenames]
sym_taxtab <- tax_table(sym)

sym2 <- phyloseq(sym_samdat, sym_otutab, sym_taxtab)

# Assign dominant symbiont clade to sample data
domclade <- sym2 %>%
  tax_glom(taxrank = "Clade") %>%                # agglomerate at clade level
  psmelt() %>%                                   # Melt to long format
  select(Sample, Abundance, Site, Species, Subtype2, Clade) %>%
  group_by(Sample) %>%
  filter(Abundance == max(Abundance)) %>%
  arrange(Sample)

sample_data(sym2)$domclade <- domclade$Clade
```

### Load bacteria data
```{r}
bact_data <- read_csv("data/STJ_OTUdata_cleaned_V_added.csv") %>%
  select(-1, -matches("W1$|W2$|W3$"), -matches("Summary")) %>%      # Remove water samples and Summary
  rename(D.labrinthiformis.OC.089 = D.labrinthiformis,              # Add missing sample name
         D.labrinthiformis.OC.136 = D.labrinthiformis.OC.084)       # Change incorrect sample name    
bact_data[1, 47] <- "dlab"                                          # Add missing species value
bact_data[3, 47] <- "OC"                                            # Add missing site name
  
# Parse sample data
bact_samdat <- bact_data %>%
  filter(between(row_number(), 1, 3)) %>%
  select(-(2:7)) %>%
  gather(sample, val, 2:ncol(.)) %>%
  spread(names(.)[1], val) %>%
  separate(sample, sep="\\.", into = c("G", "sp", "site", "id")) %>%
  unite(site, id, col="Sample", sep = "") %>%
  select(Sample, Location, Species, Type) %>%
  arrange(Sample) %>%
  mutate(HH = case_when(Location %in% c("CB", "H", "LT") ~ "outHH",
                        Location %in% c("OC", "WC") ~ "inHH")) %>%
  data.frame(., row.names = .$Sample)

# Parse OTU table
bact_otutab <- bact_data %>%
  filter(row_number() > 3) %>%
  select(1, 8:ncol(.)) %>%
  type_convert() %>%
  column_to_rownames("OTU_Name_1") %>%
  data.frame()
colnames(bact_otutab) <- rownames(bact_samdat)

# Parse taxonomy table
bact_taxtab <- bact_data %>%
  filter(row_number() > 3) %>%
  select(1:7) %>%
  mutate(tax = map(`16rRNA`, parse_taxonomy_qiime),
         tax2= map(tax, bind_rows)) %>%
  unnest(tax2) %>%
  select(1, 10:15, 9, 3:5) %>%
  column_to_rownames("OTU_Name_1") %>%
  data.frame()
  
# Create phyloseq object
bsd <- sample_data(data.frame(bact_samdat, row.names=bact_samdat$Sample))
bot <- otu_table(bact_otutab, taxa_are_rows = TRUE)
btt <- tax_table(as.matrix(bact_taxtab))
bac <- phyloseq(bsd, bot, btt)

# Subset bacteria data for samples for which we have Symbiodinium data
bac <- subset_samples(bac, sample_names(bac) %in% sample_names(sym2))
# Add dominant Symbiodinium clade to sample data
sample_data(bac)$domclade <- domclade$Clade

# Filter low abundance taxa
bac <- filter_taxa(bac, function(x) max(x) >=3 , TRUE)  # Only keep if observed >= 3x in 1 sample

# Rarefy OTU table
set.seed(1234)
bac <- rarefy_even_depth(bac)

# Transform to relative abundance
bac.p <- transform_sample_counts(bac, function(x) x/sum(x))
bac.t <- transform_sample_counts(bac, function(x) sqrt(x/sum(x)))


```

----------

# *Diploria labyrinthiformis*
```{r dlab}
# Subset D. labyrinthiformis data
dlab.sym <- subset_samples(sym2, Species == "Dlab")
dlab.bac <- subset_samples(bac.p, Species == "dlab")        # Use proportion data
dlab.bac <- prune_taxa(taxa_sums(dlab.bac) != 0, dlab.bac)  # Prune taxa not present in subset
dlab.bac <- subset_samples(dlab.bac, sample_names(dlab.bac) %in% sample_names(dlab.sym))
```

### Bacterial community structure
```{r dlab_barplots, fig.width = 10, fig.height = 4}
# Glom OTUs to Phylum
dlab.bac.phylum <- tax_glom(dlab.bac, taxrank = "Phylum")

# Plot bacterial communities by location
plot_bar(physeq = dlab.bac.phylum, fill = "Phylum", title = "D. labyrinthiformis") +
  facet_grid(~ Location, scales = "free_x") +
  theme(axis.text.x = element_text(vjust = 0.5))

# Plot bacterial communities by dominant Symbiodinium clade
plot_bar(physeq = dlab.bac.phylum, fill = "Phylum", title = "D. labyrinthiformis") +
  facet_grid(~ domclade, scales = "free_x") +
  theme(axis.text.x = element_text(vjust = 0.5))
```

### Ordination analysis (NMDS)
```{r dlab_nmds, fig.height = 3, fig.width = 4}
# Calculate Bray-Curtis dissimilarity matrix
dlab.nmds <- ordinate(dlab.bac, method = "NMDS", distance = "bray", trace = FALSE)
# Plot ordination visualizing differences among all sites
plot_ordination(dlab.bac, dlab.nmds, type = "samples", shape = "domclade", color = "Location")
# Plot ordination visualizing differences in Hurricane Hole
plot_ordination(dlab.bac, dlab.nmds, type = "samples", shape = "domclade", color = "HH")
```

### Influence of location and *Symbiodinium* (PERMANOVA)
Are location and dominant *Symbiodinium* clade significant predictors of bacterial community structure?
```{r dlab_permanova}
# Test all locations
dlab.perm <- vegan::adonis(distance(dlab.bac, "bray") ~ Location * domclade, 
                           data=as(sample_data(dlab.bac), "data.frame"), 
                           permutations=9999)
dlab.perm  
```
There is a strong effect of location, but no effect of dominant symbiont clade.

### Bacteria and *Symbiodinium* correlation (Mantel)
```{r dlab_mantel}
# Compare Bray-Curtis dissimilarities for bacteria and Symbiodinium
dlab.sym.dist <- phyloseq::distance(dlab.sym, "bray")
dlab.bac.dist <- phyloseq::distance(dlab.bac, "bray")
ecodist::mantel(dlab.bac.dist ~ dlab.sym.dist)
```
No statistical evidence that these communities are correlated in *D. labyrinthiformis*.

----------

# *Orbicella annularis*
```{r oann}
# Subset D. labyrinthiformis data
oann.sym <- subset_samples(sym2, Species == "Oann")
oann.bac <- subset_samples(bac.p, Species == "oann")  # Use proportion data
oann.bac <- subset_samples(oann.bac, sample_names(oann.bac) %in% sample_names(oann.sym))
```

### Bacterial community structure
```{r oann_barplots, fig.width = 10, fig.height = 4}
# Glom OTUs to Phylum
oann.bac.phylum <- tax_glom(oann.bac, taxrank = "Phylum")

# Plot bacterial communities by location
plot_bar(physeq = oann.bac.phylum, fill = "Phylum", title = "O. annularis") +
  facet_grid(~ Location, scales = "free_x") +
  theme(axis.text.x = element_text(vjust = 0.5))

# Plot bacterial communities by dominant Symbiodinium clade
plot_bar(physeq = oann.bac.phylum, fill = "Phylum", title = "O. annularis") +
  facet_grid(~ domclade, scales = "free_x") +
  theme(axis.text.x = element_text(vjust = 0.5))
```

### Ordination analysis (NMDS)
```{r oann_nmds, fig.height = 3, fig.width = 4}
# Calculate Bray-Curtis dissimilarity matrix
oann.nmds <- ordinate(oann.bac, method = "NMDS", distance = "bray", trace = FALSE)
# Plot ordination visualizing differences among all sites
plot_ordination(oann.bac, oann.nmds, type = "samples", shape = "domclade", color = "Location")
# Plot ordination visualizing differences in Hurricane Hole
plot_ordination(oann.bac, oann.nmds, type = "samples", shape = "domclade", color = "HH")
```

### Influence of location and *Symbiodinium* (PERMANOVA)
Are location and dominant *Symbiodinium* clade significant predictors of bacterial community structure?
```{r oann_permanova}
oann.perm <- vegan::adonis(distance(oann.bac, "bray") ~ Location * domclade, 
                           data=as(sample_data(oann.bac), "data.frame"), 
                           permutations=9999)
oann.perm  
```
There is a strong effect of location, but no effect of dominant symbiont clade.

### Bacteria and *Symbiodinium* correlation (Mantel)
```{r oann_mantel}
oann.sym.dist <- phyloseq::distance(oann.sym, "bray")
oann.bac.dist <- phyloseq::distance(oann.bac, "bray")
ecodist::mantel(oann.bac.dist ~ oann.sym.dist)
```
Highly significant correlation between bacterial and *Symbiodinium* community dissimilarity matrices in *O. annularis*. This means that corals with more similar *Symbiodinium* communities also have more similar bacterial communities, i.e., these communities covary in some way. This could be due to an effect of *Symbiodinium* in determining bacterial community composition (or vice versa), or, an effect of location driving changes in both communities.

----------

# *Colpophyllia natans*
```{r cnat}
# Subset D. labyrinthiformis data
cnat.sym <- subset_samples(sym2, Species == "Cnat")
cnat.bac <- subset_samples(bac.p, Species == "cnat")  # Use proportion data
cnat.bac <- subset_samples(cnat.bac, sample_names(cnat.bac) %in% sample_names(cnat.sym))
```

### Bacterial community structure
```{r cnat_barplots, fig.width = 10, fig.height = 4}
# Glom OTUs to Phylum
cnat.bac.phylum <- tax_glom(cnat.bac, taxrank = "Phylum")

# Plot bacterial communities by location
plot_bar(physeq = cnat.bac.phylum, fill = "Phylum", title = "C. natans") +
  facet_grid(~ Location, scales = "free_x") +
  theme(axis.text.x = element_text(vjust = 0.5))

# Plot bacterial communities by dominant Symbiodinium clade
plot_bar(physeq = cnat.bac.phylum, fill = "Phylum", title = "C. natans") +
  facet_grid(~ domclade, scales = "free_x") +
  theme(axis.text.x = element_text(vjust = 0.5))
```

### Visualize differences among samples (NMDS)
```{r cnat_nmds, fig.height = 3, fig.width = 4}
# Calculate Bray-Curtis dissimilarity matrix
cnat.nmds <- ordinate(cnat.bac, method = "NMDS", distance = "bray", trace = FALSE)
# Plot ordination visualizing differences among all sites
plot_ordination(cnat.bac, cnat.nmds, type = "samples", shape = "domclade", color = "Location")
# Plot ordination visualizing differences in Hurricane Hole
plot_ordination(cnat.bac, cnat.nmds, type = "samples", shape = "domclade", color = "HH")
```

### Influence of location and *Symbiodinium* (PERMANOVA)
Are location and dominant *Symbiodinium* clade significant predictors of bacterial community structure?
```{r cnat_permanova}
cnat.perm <- vegan::adonis(distance(cnat.bac, "bray") ~ Location * domclade, 
                           data=as(sample_data(cnat.bac), "data.frame"), 
                           permutations=9999)
cnat.perm  
```
There is a strong effect of location, but no effect of dominant symbiont clade.

### Bacteria and *Symbiodinium* correlation (Mantel)
```{r cnat_mantel}
cnat.sym.dist <- phyloseq::distance(cnat.sym, "bray")
cnat.bac.dist <- phyloseq::distance(cnat.bac, "bray")
ecodist::mantel(cnat.bac.dist ~ cnat.sym.dist)
```
No statistical evidence that these communities are correlated in *C. natans*.
